Twenty years of dynamic occupancy models: a review of applications and a look towards the future

### 2024-08-10 Manuscript/domReview.qmd
updates to manuscript and figures

---
title: "Twenty years of dynamic occupancy modelling --- what’s been done, and where to next?"
date: today
authors:
  - name: Saoirse Kelleher
    affiliation: University of Melbourne
    email: saoirse.kelleher@student.unimelb.edu.au
    orcid: 0000-0002-4614-5753
    corresponding: true
  - name: Natalie Briscoe
    affiliation: University of Melbourne
  - name: Gurutzeta Guillera-Arroita
    affiliation: Pyrenean Institute of Ecology, Spanish National Research Council
  - name: Jane Elith
    affiliation: University of Melbourne

bibliography: references.bib
embed-resources: true
---

@@

Since its introduction two decades ago by @mackenzie2003, the dynamic multi-season occupancy model has become an important tool for analysis of species occupancy throughout space and time. Dynamic occupancy models (DOMs) represent a happy medium in the ecological modelling landscape. In explicitly estimating the processes which guide occupancy (i.e., colonisation and extinction) this model class offers more advanced insights than simpler static models of species distribution which have noted deficits in their ability to describe non-equilibrium states **\[cite\]**. The capacity to account for imperfect detection provides a strong additional benefit given the near-ubiquity of this source of bias in ecological field data. Even with these advantages, data input requirements for DOMs remain achievable for many research programs, particularly compared to other more complex process-explicit models.

Potential use cases of this model class are wide-ranging. Simple estimates of occupancy through time have value in many monitoring programs, particularly for threatened or invasive species where distributions are likely to exist in a non-equilibrium state **\[cite?\]**. These models can also be used in hypothesis testing of specific relationships, with examples including the effect of fire on extinction probability and the influence of species interactions on occupancy **\[cite,cite\]**. A less frequently pursued yet promising application of DOMs lies in their ability to make predictions, either to unsurveyed areas or into the future [@briscoe2021; @kéry2013a].

This review aims to provide an overview of the first two decades of the dynamic occupancy models, including its development and how authors have implemented it in their investigations of ecological questions. We present key guidelines for authors seeking to build these models in their work, and outline the key priorities and opportunities for future research incorporating this flexible and important model class.

### The dynamic occupancy model

Dynamic occupancy models jointly describe two processes: the ecological process of site occupancy describing the presence or absence of a species at a given site in a given point in time; and the observational process of detection describing whether a species is observed given that a site is occupied. The basic model form as described by @mackenzie2003 is depicted in @fig-modelform. Under this parameterisation sites exist in either occupied or unoccupied states which they may transition between through time. Occupancy state in the first time-step is given by the initial occupancy parameter, and in ensuing time-steps occupancy is described as a Markovian process governed by probabilities of colonisation and extinction. Given that a site is occupied, the probability of observing the species during a survey is given by the detection parameter. Under this formulation, it is assumed that false-positive detections do not occur.

![The original parameterisation of the dynamic occupancy model. The ecological process of occupancy is guided by the initial occupancy, colonisation, and extinction parameters; the observational process is guided only by the detection parameter.](Figures/ModelForm.png){#fig-modelform width="1000"}

To disentangle the ecological and observation processes and estimate the detection parameter, the DOM relies on a hierarchical sampling design shown in @fig-surveys. Under this design, observations at a site occur during distinct, time-bound seasons in which sites are considered 'closed' to changes in occupancy. Within seasons multiple observations are conducted, which permit estimation of the probability of detection conditional on occupancy. These repeat observations may be conducted by a variety of methods. Most frequently surveys are simply repeated on separate occasions, although alternative methods exist. These include surveys conducted at multiple locations within a site **\[cite\]**, use of multiple observers during a survey **\[cite\]**, or by recording time elapsed before a detection is recorded **\[cite\]**.

![The sampling design of the dynamic occupancy model.](Figures/SurveyDesign.png){#fig-surveys fig-align="center" width="589"}

Dynamic occupancy models require a few assumptions with important implications for users.

1.  **False positive detections do not occur.** While this assumption can be safely met in some studies, it is not guaranteed when working with more cryptic species. Several authors have commented on the perils of not accounting for false positives where they occur (@miller2015, @mcclintock2010). Furthermore, @berigan2019 notes that in some cases genuine identifications of a species may be considered false positives if they do not represent the right kind of occupancy - in this case, transient individuals where the authors were interested in breeding occupany. Extensions to the dynamic occupancy model which account for false positive error are available (@royle2006, @miller2011, @miller2013) and should be carefully considered where this assumption cannot be met.

2.  **Sites are closed to changes in occupancy between seasons.** This assumption, best known as the 'closure assumption,' has also been subject to considerable discussion on the bias which is introduced when it is violated (@rota2009, @otto2013). Closure is dependent not only on the behaviour of the species, but also the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discusses the difficultly in distinguishing temporary emigration and local extinction. Model extensions to relax the closure assumption have been developed, including @kendall2013 's approach using staggered arrival and depature periods between sites. More pertinent for most studies, however, is a careful consideration of ecologically and observationally relevant seasons which will allow the estimation of an appropriate 'type' of occupancy, whether that be shorter or longer term.

3.  **No modelled heterogeneity in ecological or observational parameters.** Relative to the former two assumptions, considerably less research has been conducted on the impact of unmodelled heterogeneity on parameter estimates. Particularly at larger spatial scales, environmental factors are near-guaranteed to influence species occupancy and observation.

# Methods

## Article solicitation

This review is focused on the dynamic, multi-season, implementation of the occupancy model of @mackenzie2003. To qualify for inclusion articles must fit a model which fulfils these criteria: i) multiple sites which can exist in at least two states, including an occupied and unoccupied states; ii) multiple survey periods, between which sites may change state conditional on the prior survey period's occupancy state and at least one transition probability; iii) contains at least one detection parameter; iv) uses real (i.e., not simulated) field ecological data.

Papers were solicited for inclusion in two stages. In a preliminary analysis, four search strings ("dynamic occupancy model," "occupancy dynamics model," "incidence function model," and "multi-season occupancy model" were fed into Google Scholar **\[cite\]**. Up to 100 articles were downloaded for each string in order of appearance. Articles were divided into four temporal strata: 2000-2005, 2006-2010, 2011-2015, and 2016-2021. For each combination of search string and year strata, either a random 25% or 5 articles (whichever was greater) were processed. Articles which did not meet inclusion criteria were replaced with another from the same search string and strata where possible.

After completion of the preliminary sample authors recognised significant uncertainty in how the model of interest was described, particularly in their first decade of use, and noted that the incidence-function model is a distinct model class **\[cite\].** To increase sample size, a complete list of articles which have cited @mackenzie2003 was downloaded from Web of Science. Articles in this list were split into the same year strata, plus an additional 2022-2023 category. 15 articles were drawn from each strata and again processed with replacement from their strata.

Details on methods for each article were recorded in detail, including the full suite of covariates considered for inclusion in modelling. The full spreadsheet of reviewed articles is available in **Appendix I**.

## Analysis

For each included article we collected data collected on authorship, stated research objectives, study taxa and field survey locations, data collection methods, covariates, and model selection and evaluation. Particular emphasis was placed on the covariates considered by authors, with no distinction between covariates which were or were not included in final models. As we were principally interested in the modelling process, our analysis did not directly explore final models or the results generated by those models. A spreadsheet of reviewed articles and collected data is provided in Appendix I.

# Results

A total of **\[number\]** articles were included in the review, **\[number\]** from the preliminary sample and **\[number\]** from the secondary sample. Based on the acceptance rate from the secondary sample, an estimated **\[hit number\]** of the **\[total number\]** articles citing @mackenzie2003 would have met inclusion criteria. Total published articles have increased considerably over time.

![Estimated coverage of articles included in this review](Figures/CoveragePlot.jpeg){#fig-coverage}

Articles were published with first-authors from a total of **\[number\]** countries, with authors from the United States producing **\[zz\]**% of articles. **\[x\]**% of articles had at least one author based at an academic institution, **\[y\]**% had authors at government institutions, and \[**z\]**% had authors from non-profit organisations.

## Study locations, focal species, and data collection

Data for DOMs were collected from study areas on all continents and in **\[xx\]** countries, with the United States again representing a majority of the total articles published. These study areas ranged in size from less than 1 to over 1 million square kilometres (@fig-StudyDetails a).

Focal species were largely vertebrate taxa, although 6 articles fit models to invertebrates. A single non-animal taxa was represented in the sample, from an article fitting a DOM to a rare species of lichen @belinchón2017. [aa%] of articles modelled threatened taxa, and \[bb%\] modelled species invasive to the study area. The majority of articles (50, X%) fit single-species DOMs to one taxa. Of the remainder, 26 (X%) fit independent single-species models to more than one taxa. 8 articles fit (X%) multi-species implementations with explicit interactions between species (@fig-StudyDetails b).

Data used to fit dynamic occupancy models included human survey data such as transects or point counts (X% of articles), live-trapping data (X%), camera traps (X%) or bioacoustic monitors (Y%), and citizen science data (Z%); some articles incorporate multiple detection methods to build their detection datasets. Study duration, defined as the time elapsed between the first and last surveys conducted, ranged from 1 month to 40 years with the median study running for X years. The number of seasons ranged from 2 to Y with a median of Z; the number of observations per season ranged from X to Y with a median of Z. The number of sites surveyed per article ranged from X% to Y%, with the median study including Z sites (@fig-StudyDetails c).

![](Figures/ReviewBox.png){#fig-StudyDetails}

## Covariates

A summary of the covariates classes considered in the review sample is presented in @tbl-covariates. The most frequent

| Covariate type | Proportion of articles considering | Proportion including on Initial Occupancy | Proportion dynamic | Proportion directly observed | Frequency of interaction terms | Frequency of polynomial terms |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |
|                |                                    |                                           |                    |                              |                                |                               |

: Covariates considered in dynamic occupancy models {#tbl-covariates}

![](Figures/CovParamPlot.jpeg){#fig-covariates}

### Complexity

![](Figures/ModelComplexity.png){#fig-complexity}

### Selection and evaluation

Bayesian models were far less likely to implement a model selection protocol compared to frequentist models: 30% of the former fit only a single model defined *a priori,* compared to 10% of the latter (@fig-selection). In frequentist models, stepwise selection protocols were the most popular choice. The vast majority of frequentist models (XX%) used AIC or AICc to identify the best model.

![](Figures/SelectionFigure.jpeg){#fig-selection}

Objective evaluation of model fit was rare throughout the sample, regardless of selection protocol. XX% of articles calculated non-relative metrics, either via in-sample validation, bootstrapped goodness-of-fit tests, or out-of-sample validation.

# Discussion

### Usage of dynamic occupancy models

Our review sample displays considerable diversity in implementations of dynamic occupancy models with widespread representation from mammals, birds, and herptiles. This assemblage contrasts with the related stochastic patch occupancy models (SPOMs, @gutiérrez-arellano), where the majority of models are fit to invertebrates, plants, and fungi. This may reflect SPOMs lack of consideration for imperfect detection and assumption of discrete habitat patches, limitations which are unlikely to be fulfilled by most vertebrate taxa.

DOMs also display great variation in scale, from local studies with a handful of sites to continental-scale studies spanning decades.

Several types of detection data have been successfully incorporated DOMs. This flexibility may be one of the more notable aspects of dynamic occupancy models; while the heirarchical data inputs may be more complex than those required by other models of species distributions, authors have been able to successfully manipulate datasets from many sources to meet requirements.

### Practices in the modelling process

Model selection approaches diverge between Bayesian and frequentist implementations of the model. While model selection will naturally look different between these two categories, it is unclear what effect a lack of model selection may have on the outputs produced by Bayesian models and it is likely that Bayesian model selection approaches should be further explored @hooten2015.

Notably, the popular sequential approach to model selection (where one sub-model is assessed at a time) may not provide optimal results as shown in @morin2020. Considerably more research is required on this topic to clarify how authors should best approach the model building process.

### Key considerations for model users

Implementation of dynamic occupancy models requires considerable thought to ensure that model outputs correspond with author expectations and ecological realities. Based upon our review of the existing literature we discuss key questions authors should carefully engage with during the modelling process.

1.  **Are seasons ecologically meaningful, and how do they affect the definition of occupancy?** As previously discussed, the duration of the season has important implications for both the reliability and proper interpretation of model results. Seasons which are not ecologically meaningful may run afoul of the closure assumption resulting in associated biases. Also relevant is how the definition of 'occupancy' may shift with season length and life-history of a species — for many species, occupancy within a year and within a week may be considerably different concepts, with shorter intervals more closely relating to the idea of 'use.' With the increasing proliferation of continuous detection methods such as camera traps and acoustic monitors where the definition of 'season' is to some degree arbitrary, it is increasingly important to thoughtfully consider how this element of the DOM is defined.

2.  **Are key drivers of occupancy and detection incorporated?** While it is effectively impossible to *fully* describe heterogeneity across a landscape, authors should make their best efforts to capture the drivers of occupancy and detection most likely to influence their study species. This is true even when another covariate is of principal interest to investigators — failure to include other major determinants of occupancy is likely to bias parameter estimates. One place where this is of particular importance is in determining drivers of detection, which is not only subject to observational conditions but also ecological determinants which may cause the species to spend more time at a location and thus be more frequently available for detection.

3.  **Ensure covariates reflect ecological complexity.** Many ecological relationships are non-linear and may require polynomials terms to be accurately represented. This is particularly true at large spatial scales where a larger portion of a species ecological niche is shown in surveyed data. Authors should question whether a linear term is sufficient to capture hypothetical relationships in their study systems and consider quadratic terms where appropriate.

4.   **Consider model evaluation.** Model evaluation remains rare in ecological research relative to other fields embracing complex modelling, likely due to a reluctance to 'throw away' data for use in testing. This concept should be challenged, particularly for larger scale long-term projects where reserving a fraction of data for testing is unlikely to substantially reduce precision of parameter estimates. For studies where this is genuinely not possible, reporting results of goodness-of-fit tests may still provide valuable information to readers.

### Priorities for future research

Work on dynamic occupancy modelling over the past 20 years has developed this model class into an important tool for describing species distributions and ecological relationships. Historically, methods-development research for dynamic occupancy models has focused on questions of model structure — namely ideas around the closure assumption and imperfect detection. This emphasis likely reflects the DOM's roots in the mark-recapture literature, where these topics have been discussed for several decades. Considerably less emphasis has been placed on the relation of model parameters to environmental factors via covariates and the model selection processes used to do so. This stands in stark contrast to related correlative species distribution models, where these topics have been the subject of considerable research over the years. There are opportunities for the DOM literature to learn from species distribution models, and more targeted research on the influence of the model selection process is necessary.

The use of DOMs in prediction has remained fairly limited despite significant potential for this use-case. Papers in our sample which have taken advantage of this potential included @mcgowan2020 who used DOMs to make predictions of black rail distributions under various management scenarios, and @saunders2020, who validated predictions from ecological niche models using continent-scale DOMs. Further implementations of these models may benefit from the aforementioned necessary research on model selection to ensure robust outputs.

A final opportunity for future research relates to autonomous detection methods, such as camera traps and bio-acoustic monitors — these technologies have permitted collection of enormous quantities of data particularly well-suited to dynamic occupancy models, but have also raised additional questions. The definition of seasons is of increased importance for these data types, where division of data into 'seasons' and 'observations' is arbitrary. Further research should explore the implications of this phenomenon and additional modelling opportunities made possible by this novel data type.

The past two decades have shown that dynamic occupancy models are an excellent tool for addressing a wide range of ecological questions relating to species occupancy throughout space and time. The flexibility afforded by model extensions and the considerable advantages of the class over other species distribution models is likely to ensure that these models continue to increase in popularity. Given their strong use on threatened taxa and excellent potential to support environemental management, it is important that research continues to support model users in creating the most reliable models possible.

>>>>

---
title: "Twenty years of dynamic occupancy models: a review of applications and a look towards the future"
date: today

authors:
  - name: Saoirse Kelleher
    affiliation: University of Melbourne
    email: saoirse.kelleher@student.unimelb.edu.au
    orcid: 0000-0002-4614-5753
    corresponding: true
  - name: Natalie Briscoe
    affiliation: University of Melbourne
  - name: Gurutzeta Guillera-Arroita
    affiliation: Pyrenean Institute of Ecology, Spanish National Research Council
  - name: Jane Elith
    affiliation: University of Melbourne

abstract: | 
  Describing patterns of species occupancy across landscapes and throughout time is a fundamental aspect of much ecological research. The dynamic multi-season occupancy model is an important tool for analysing wildlife occurence data -- by explicitly estimating colonisation and extinction probabilities while accounting for imperfect detection, it strikes a favourable balance between realism and feasibility. These models can be used in similar ways to more conventional species distribution models, which can struggle to deal with imperfect detection and non-equilibrium populations.

  We present a review of applications of the dynamic occupancy model in the twenty years since its initial publication. Our findings indicate that this model demonstrates exceptional flexibility with suitability to address a wide range of ecological questions, and that it may have  broader potential than commonly thought. We also note that practices in fitting dynamic occupancy models are highly variable, and that the existing literature lacks clear answers for best practices. This contrasts with the strong focus on modelling practices in the species distribution model literature, which we suggest may be a valuable resource for improving how authors implement dynamic occupancy models.
---

@@

Ecologists have long sought to describe the processes by which species are distributed across landscapes and throughout time [@humboldt1849], and an understanding of where a species exists, which factors lead to its presence, and how its distribution may change is increasingly important in a dynamic and changing world. Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs assessing conservation status, and identifying potential drivers of occurrence can help inform potential management actions. Robust knowledge of the occupancy patterns of a species also helps us to predict where a species is most likely to occur, both under present conditions and hypothetical future scenarios.

The inherent value in understanding these patterns of occurrence has led to the development of several techniques for modelling occupancy. However, several factors can make occupancy challenging to describe quantitatively. For instance, when conducting surveys it is often impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer. This 'imperfect detection' is a near-ubiquitous characteristic of wildlife presence/absence data and is known to bias estimates of occupancy in conventional SDMs [@lahoz-monfort2014; @gu2004]. Despite this, many popular models make no adjustments for this source of bias [@kellner2014]. Another challenge lies in modelling occupancy under non-equilibrium conditions where a species' distribution or relationship to its environment is in flux, as occurs during range shifts or biological invasions. Conventional correlative SDMs often perform poorly in these scenarios [@dormann2007; @elith2010], a noteworthy challenge given the increased frequency of these events in the Anthopocene [@bertelsmeier2013; @lenoir2015].

Occupancy models, first introduced in @mackenzie2002, offer an alternative framework for modelling occurrence while accounting for these challenges. By using repeat visits to each site, static occupancy models explicitly describe the observation process and account for imperfect detection to correct for bias in estimates of site occupancy. @mackenzie2003 extended this model to account for changes in occupancy across time periods by explicitly describing processes of colonisation and extinction, relaxing the assumption of equilibrium present in other models of species occurrence. This dynamic occupancy model (DOM) is well-suited to address many research questions in ecology, offering a balance between achievable input data requirements and the complexity necessary to describe intricate natural systems.

This review aims to provide an overview of dynamic occupancy models, their uses, and the ways that they are implemented. Following an introduction to the model form and assumptions, we present a systematic review exploring how authors have used the DOM in the two decades since their development, with emphasis on how they collected data, selected covariates, and evaluated their models. Based on these results we highlight the DOMs flexibility as a tool for understanding occupancy, examine how authors have approached the modelling process, and outline key priorities for future research involving this important model class.

## Dynamic occupancy model form and assumptions

Dynamic occupancy models unite two processes: the ecological process of site occupancy describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed given that a site is occupied (@fig-modelform). Under this parameterisation, sites exist in either occupied or unoccupied states. In the first step-step, occupancy state at each site is determined by the probability of initial occupancy.In subsequent time steps occupancy state is described as a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation and extinction. The observation component of the DOM accounts for imperfect detection - at occupied sites, the detection probability describes whether or not the species is observed during a survey. Under the DOM's standard parameterisation, it is assumed that false-positive detections never occur at unoccupied sites.

![The dynamic occupancy model as described by @mackenzie2003. In the ecological process submodel, the occupancy state of any given site is determined by the *initial occupancy* parameter. In subsequent time-steps, unoccupied sites may become occupied according to the *colonisation probability* and occupied sites may become unoccupied according to the *extinction probability*. The observational process submodel accounts for imperfect detection: during a survey at occupied site, the species may be detected or not according to the *detection probability*. At sites which are truly unoccupied, is it assumed that the species is never detected.](Figures/ModelForm.png){#fig-modelform}

To disentangle the ecological and observational processes the DOM requires a hierarchical sampling design as depicted in @fig-surveys. Under this design, observations at a site occur during distinct, time-bound seasons within which sites are considered closed to changes in occupancy. In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy. Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@mackenzie2005].

![The sampling design of the standard dynamic occupancy model. During seasons, also called primary occasions, sites are considered closed to changes in true occupancy state; occupancy state may only change between seasons. Within each season, multiple observations ('secondary occasions') are conducted to record the observed presence or absence of the species at each site. These multiple observations may be recorded in many ways: sites can be revisited multiple times within a season, surveys can be conducted at multiple points within a larger site, multiple observers can conduct surveys contemporaneously, or the time elapsed prior to a detection occurring can be recorded. It is not necessary for the same number of observations to occur in each year or each site, allowing for greater flexibility in data inputs.](Figures/SurveyDesign.png){#fig-surveys fig-align="center"}

The descriptions above are based on the original form of the dynamic occupancy model as prescribed in @mackenzie2003. Numerous model extensions and alternative formulations are also available, including implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models[@dorazio2010]. For a comprehensive discussion of the most common extensions and their applications see @bailey2014, as well as @devarajan2020 for a review of multi-species occupancy models.

Dynamic occupancy models make a small number of assumptions with important implications:

I.  **False positive detections do not occur**. While this assumption can be safely met in many studies, it is not necessarily guaranteed when working with more cryptic species or less reliable survey methods. @mcclintock2010 and @miller2015 comment on the bias induced when false positives occur and are not accounted for, highlighting the need for authors to consider how certain their detections truly are. Significantly, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019]. Where this assumption can not reasonably met, aforementioned extensions designed to account for false positive error should be considered.

II. **Sites are closed to occupancy between seasons.** This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion around the bias which is introduced when it is violated [@rota2009; @otto2013]. Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discuss the difficultly in distinguishing between temporary emigration and local extinction. Model extensions to relax the closure assumption have been developed, including @kendall2013's approach using staggered arrival and departure periods between sites. A more pertinent approach for most studies, however, is careful consideration of ecologically relevant seasons corresponding to an appropriate definition of occupancy.

III. **Heterogeneity in occupancy and detection is accounted for.** As with any approach for modelling species distributions, it is assumed that DOMs appropriately capture variation in occupancy patterns and species detectability across the study system. Generally, this is achieved by allowing model parameters to vary with respect to covariates representing environmental factors which may be expected to influence these characteristics. No model will fully account for the complexity inherent in patterns of species occupancy; however, failure to capture key drivers can introduce bias. Compared to the first two assumptions, this aspect of DOMs has been less thoroughly discussed and comparatively little is known how this latent heterogeneity can influence model estimates.

## Building dynamic occupancy models

A key aim of this review is to investigate how authors have approached the steps of the modelling building process, and how these practices have changed since the introduction of these models 20 years ago. The process of fitting DOMs requires decisions by model builders at several key points with meaningful implications for how outputs should be interpreted. This process begins with the project's inception and plans for data collection. The definition of occupancy is contingent on

Authors must also consider the temporal aspect of occupancy by defining primary and secondary occasions. In many cases, this is a simple decision driven by data collection - take a typical bird survey program, where sites may be visited on a few occasions in the breeding season of each year; years and individual visits within years will generally be used as primary and secondary occasions, respectively. However, this choice may not be as clear with other data collection methods. Citizen science programs, camera traps, acoustic monitors can all yield somewhat continuous data which must be arbitrarily segregated into primary and secondary occasions. In this context, it is important to consider both the temporal scale of interest for occupancy and whether or not closure can reasonably be assumed between primary occasions.

Once data has been coerced into a suitable structure for fitting DOMs, modellers must also consider which environmental factors to incorporate into the model, as well as how to incorporate them. These factors will vary considerably by study, but should generally be plausible drivers of variation in either occupancy dynamics or survey observability. Determining which candidate covariates should be included in a final model for drawing inference will generally involve some form of model selection. Covariate selection can be particularly challenging for DOMs relative to other models like SDMs due to their increased complexity. Consider a model with 6 candidate covariates, each of which can be included in any parameter. The number of permutations of this model is given as (2^n^)^p^, where n is the number of candidate covariates and p the number of parameters. A simple single-parameter SDM would have 64 possible formulations, whereas a typical 4-parameter DOM would have over 16 million candidate models given the same covariate pool. For many studies, this precludes exhaustive comparisons of candidate models and demands some form of simplified model selection protocol [@doherty2012]. Multi-model inference is also a common strategy with DOMs, allowing for uncertainty in the difference in fit by averaging estimates from across competing models [@burnham2004].

Several computational tools exist for implementing the DOM and related models. Widely-cited frequentist options include the R package 'unmarked' and the GUI-based programs PRESENCE and MARK. Bayesian DOMs can be written in programs such as JAGS, BUGS, or STAN and fit via R packages such as rJAGS or brms*.* The recently developed R package 'ubms' offers a straightforward Bayesian implementation through an interface similar to *unmarked,* without requiring writing the DOM from scratch.

Upon selecting a final model, authors may evaluate model performance by different methods dependent on the objectives of their studies. While no dedicated goodness-of-fit test exists for DOMs, a version is available for single-season models [@mackenzie2004] which has been extended to multiple seasons; functions are available in *unmarked* and *aicCmodavg.* Another possibility for assessing model fit and predictive ability is via validation. As true occupancy state is generally not known for DOM data, the typical response variable is the observed occupancy state.

# Systematic review methods

To assess how DOMs have been applied in the years since their introduction, we gathered a representative sample of articles fitting them to field ecological data. A pool of candidate articles was generated using two queries on Web of Science. The first included all articles from 2004-2023 citing @mackenzie2003. The second query searched articles matching "dynamic occupancy model\*", "multi-season occupancy model\*", and "occupancy dynamic\*" in any field, articles with each of "occupancy", "colonization", "extinction/persistence", and "detection" in any field, and articles with the term "occupancy" near "dynamic".

As we were interested in how DOMs use has changed through time, we divided articles across five four-year strata: 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024. From each of these strata we randomly selected 20 articles for inclusions in the review; articles which did not meet inclusion criteria were replaced from within their own strata.

Our review is focused on applications of the dynamic, multi-season occupancy model of @mackenzie2003 and its extensions. To reflect this, we included articles which fit a model meeting the following criteria:

I.  Uses non-simulated, field-collected presence-absence data.

II. Describes multiple sites which can exist in at least two states including *occupied* and *unoccupied.*

III. Has multiple seasons, between which sites may change states conditional on the prior season's occupancy state and transition probabilities such as colonisation and extinction.

IV. Contains at least one parameter describing the detection process.

For each article we recorded key details on authorship, research objectives, study taxa and system, survey methods, and modelling approach. To assess patterns of authorship, the location of each article's first author's primary affiliation was recorded. *All* affiliations for all authors were classified into one of four categories: academic institutions, governmental organisations and institutes, non-governmental organisations, and private sector companies.

We recorded details on the type of taxa (bird, mammal, etc.) modelled in each article, and how multiple species were modelled where applicable. Taxa were denoted as threatened either when they are listed on the IUCN Red List of Species or when authors otherwise indicate that they are threatened. This deference to authors' representation of conservation status was made to account for sub-species which lack listings or species which are of more local concern. Study locality and size was documented; the size of the study area being defined as the intended area of inference containing all sites, recorded as an order of magnitude to partially account for uncertainty in reporting.

To examine the reasons why authors used DOMs, we allocated each article to one or more category of objective based on the study's stated aims. The possible categories were *Estimating parameters*, where authors express interest in estimates of site occupancy, colonisation, extinction, or detection probabilities; *Testing hypotheses*, where authors explore specific predefined relationships between covariates and model parameters; *Identifying drivers*, where authors attempt to find which covariates influence model parameters; *Generating predictions*, where authors estimate model parameters either for unsurveyed sites or under possible future conditions; and *Methods development*, where authors introduce, test, or demonstrate aspects of dynamic occupancy models.

Particular emphasis was placed on describing how authors conducted their modelling from covariate selection through model evaluation. We recorded all covariates considered in each study, regardless of whether they were or were not included in final models. Key traits of each covariate were recorded including their general category, whether they were directly observed or remotely sensed, whether they were static or varied between seasons, and how they were represented in the model (as a linear term, a polynomial term, or as part of an interaction with another covariate). Model selection procedures were sorted into non-exclusive categories including *A priori*, where only one model was considered; *candidate suite*, where a predefined set of models was considered; *procedural*, where covariates were selected parameter-by-parameter; *exhaustive*, where all possible model combinations were fit; *simple precursors*, where selection was preceded by another simpler model implementation; and *other*. Model evaluation was similarly categorised - options in this case included *goodness of fit*, where absolute metrics of fit were used; and *out-of-sample validation* and *in-sample-validation*, where predictive performance was evaluated using in-sample or out-of-sample data respectively. For the full spreadsheet of data collected and further details on categorisation, see **Appendix I.**

A total of 84 articles were included in this review. Based on the acceptance rate from the secondary sample, an estimated 448 of the 1066 articles in our sample would have met inclusion criteria (@fig-coverage).

![Estimated coverage of articles included in this review, based on the secondary sample's rate of inclusion for each strata.](Figures/CoveragePlot.jpeg){#fig-coverage}

# Applications of DOMs

Dynamic occupancy models are considerably flexible in many ways: they may used to achieve several different objectives, fit to presence-absence data from a variety of sources, and applied to studies at a wide range of spatial and temporal scales. Study locations included in our sample are globally distributed [@fig-StudyDetails A] and span a diverse range of ecosystems. Notably, these study areas vary considerably in size — the smallest study locality included in our sample studied insect occurrence in a rainforest plot smaller than one square kilometre [@basset2023], while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@clement2019].

![A) Study areas: Data for a majority of studies were collected from study locations in the United States. The size of study areas was log-normally distributed, with the median study area falling between 1000 and 10000 square kilometres. \n B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles. 50 articles fit models to only a single taxa, 26 fit independent models to multiple taxa, and 8 fit multi-species models with explicit interactions between taxa. \n C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as more modern methods like camera traps and acoustic monitors. 10 articles incorporated data from citizen science projects. Project scale was also variable, with the median study running for 3.25 years and covering 98 sites.](Figures/ReviewBox.png){#fig-StudyDetails}

### Focal systems

Although studies were conducted in 21 countries and on all continents except for Antarctica, the majority of reviewed articles used data collected in the United States of America. This strong geographic trend starkly contrasts with SDMs far more international reach; @araújo2019's review of SDM applications found that no more than 20% of studies came from any single continent. The reasons behind this disparity are unclear, but a possible explanation could relate to the DOM's roots in the mark-recapture literature and a history of publication in American journals. These differences in where users conduct their research may also contribute to the seemingly limited overlap between these models' respective bodies of literature.

DOMs have more rarely been applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community. However, there are exceptions — @belinchón2017 fits a DOM to lichen data, and @cook2022 uses them to model the spread of chronic wasting disease. The latter's application to disease dynamics is not unique, and DOMs have been a reasonable choice for modelling disease dynamics. @mores2020 and @padilla-torres2013 have each used DOMs to model mosquito dynamics due to their importance as a human disease vector, and @bailey2014 discusses the feasibility of applying the DOM to study the spread of chytrid fungus in amphibian habitat.

Differences in use between SDMs and DOMs are also evident in the species targeted for modelling. Where SDMs show broader coverage of plants, invertebrates, and aquatic species [@araújo2019], 93% of studies in our sample of DOMs modelled terrestrial vertebrate fauna with @belinchón2017's study on lichen the sole non-animal taxa representative. These differences are most easily explained by the DOM's emphasis on imperfect detection, a concern which is less prioritised (though not non-existent [@chen2013]) in studies on non-mobile species like plants. We find DOMs have been frequently used to model threatened taxa (32% of studies) and invasive species (11% of studies), taxa for which they may be expected to excel due to their more explicit handling of changes in occupancy. One group for which DOMs have only rarely been used is aquatic species, as in @fisher2014's application to invasive salmon or @falke2012's model on Great Plains stream fishes.

While most articles focus on a single taxa, 40% of the studies in our sample fit models to more than one taxa ('taxa' defined here as one or more species modelled as one). Of these, 75% fit independent DOMs to multiple species, and 25% fit explicitly multi-species implementations. These multi-species models varied in structure: some use species-specific effects to describe metapopulations by including hundreds of species in a single model [@dorazio2010; @hendershot2020], while others fit models which explicly account for species interactions and estimate conditional occupancy, colonisation, extinction, and detection [@lesmeister2015; @fidino2019]. While they did not fit multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012]. Working with large numbers of taxa does raise additional questions, as the level of complexity which can realistically be applied to each individual taxa is likely to be reduced for practicality's sake. With multi-species community frameworks, environmental covariates must be relatively general; and when fitting large numbers of independent models covariate selection processes are likely to be more restricted due to computational limitations.

### Data collection and inputs

DOMs show exceptional diversity in the scales at which they can be implemented. The smallest study area in our sample was less than one kilometre squared, and the largest over one million. Study duration shows similar variation — data collection ran for for less than a month in the shortest study, and for 40 years in the longest, with a median duration of 3.25 years. This reiterates the flexibility of dynamic occupancy models, with explorations of hyper-local patterns of occupancy and analyses of continental-scale distribution using the same underlying model structure. The detection data collected from these surveys came from a variety of detection methods: 69% of articles used human observations, 20% conducted live trapping, 12% used camera traps, and 1% used bioacoustic monitors (@fig-StudyDetails).

The flexibility of DOM inputs data allows for creative uses of detection data from a range of sources. @marescot2020 models a relationship between wildlife and human poachers, using ranger-reported signs of poacher activity to create detection histories. In a rare example of a DOM applied to a marine species, @pendleton2022 uses aerial transects broken up into grid cells to study the occupancy dynamics of whales in the Northeastern United States.
